Manuscript Figures

### 2021-09-07 src/manuscript_figures.Rmd
Update luad vs lusc analysis figures - drop panel E and F (feature contribution to top LFs, cancersea gene set score distributions) - add an analysis of top omics markers associated to top LFs - draw decision tree + heatmap of top markers in panel E

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = params$workdir)
settings <- yaml::read_yaml(params$settings_file)
message(settings$GDCdataDir)
FigureFolder <- params$figures_folder
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
library(survminer)
library(survival)
library(caret)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'

@@

ens2hgnc <- readRDS(settings$ens2hgnc)

@@

```{r}
get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}
```

```{r pancancer_prediction, fig.height=9, fig.width=12}

# In this section, we make figures about pan-cancer classification 
# and some benchmarks of cohort-based unsupervised clustering across cancers

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 8.27, height = 11.69, units = 'in')
Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type. B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs."

@@

```{r}
# Analysis of non-small-cell lung cancer types (LUSC vs LUAD)
# reference study: https://www.nature.com/articles/s41598-020-77284-8
M <- LFs$maui_factors$nsclc.cnv_gex_meth_mut
M <- scale(M) #remove_redundant_variables(M, perc = 99)
ct <- surv[match(rownames(M), bcr_patient_barcode)]$project

# find type-specific LFs
dt <- get_factor_specific_variables(M, ct)
top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 1)

@@

# top lf for LUSC
top_lf_lusc <- top_vars[ref_cl == 'TCGA-LUSC']$variable
top_lf_luad <- top_vars[ref_cl == 'TCGA-LUAD']$variable
df <- cbind(df, 
            M[,top_lf_lusc,drop = F],
            M[,top_lf_luad,drop = F])
p <- ggplot(df, aes(x = tSNE1, y = tSNE2)) 
p2 <- p + geom_point(aes(color = group)) + 
  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
            aes(x,y,label = text.label), size = 5)
p3 <- p + geom_point(aes_string(color = top_lf_lusc)) + 
  scale_color_gradient(low = 'gray', high = 'red') +
  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
            aes(x,y,label = text.label), size = 5)
p4 <- p + geom_point(aes_string(color = top_lf_luad)) + 
  scale_color_gradient(low = 'gray', high = 'red') +
  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
            aes(x,y,label = text.label), size = 5)
p5 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 'nsclc', 'cnv_gex_meth_mut', 
                                     factors = ct, top_lfs = 1, top_features = 500) + 
  labs(y = 'Relative Contribution\nto top latent factor')

dt <- data.table(M[,top_vars[['variable']]], keep.rownames = T)
dt <- dt[rowSums(dt[,2:3]) > 0] 
dt$pr <- surv[match(dt$rn, bcr_patient_barcode)]$project

# alternatively, split samples by LF scores, plot gene set scores 
scores <- import_geneset_scores(geneSetScoresFolder, annotation = 'cancersea')
S <- data.table(scores[intersect(rownames(scores), rownames(M)),], keep.rownames = T)

dt <- merge(dt, S, by = 'rn')

mdt <- melt(dt, measure.vars = setdiff(colnames(S), 'rn'))

p6 <- ggplot(mdt, aes(x = value, y = variable)) + 
  ggridges::geom_density_ridges(aes(fill = get(top_lf_lusc) > get(top_lf_luad)), alpha = 0.7) +
  labs(x = 'CancerSEA gene set scores', y = 'Cancer Cell States') + 
  guides(fill=guide_legend(title=paste0(top_lf_lusc, " > ",top_lf_luad)))

Fig <-  cowplot::plot_grid(p1, p2, p3, p4, p5, p6, 
                         labels = 'AUTO', ncol = 2)

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.luad_vs_lusc.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       Fig, units = 'in', 
       width = 10, height = 12)
Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI latent factors.
A) ROC curves with AUC values on testing data (40% of the dataset) using an Elastic Net model with 5-fold cross-validation. B) t-SNE plot of MAUI LFs colored by cancer types. C) t-SNE plot of MAUI LFs colored by 
the top most-predictive latent factor for TCGA-LUSC samples. D) t-SNE plot of MAUI LFs colored by the top 
most predictive latent factor for TCGA-LUAD samples. E) Percentage of top 500 contributing features each of the top most predictive latent factors broken down into categories by cancer type and data type. F) Cancer Cell State score distribution of samples grouped by whether the sample has a higher score for LUSC-predictive latent factor than the LUAD-predictive latent factor."
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = params$workdir)
settings <- yaml::read_yaml(params$settings_file)
message(settings$GDCdataDir)
FigureFolder <- params$figures_folder
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
library(survminer)
library(survival)
library(caret)
library(treeheatr)
library(partykit)

theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'

@@

ens2hgnc <- readRDS(settings$ens2hgnc)
cpg2hgnc <- readRDS(settings$cpg2hgnc)

@@

```{r}
get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}

map_ids <- function(markers, cpg2hgnc, ens2hgnc) {
  dt <- data.table::data.table('id' = gsub("^.+?\\.", "", markers), 
                               'type' = sub("(^.+?)\\..+$", "\\1", markers), 
                               'variable' = markers)
  do.call(rbind, lapply(names(split(dt, dt$type)), function(x) {
    message(x)
    if(x == 'meth') {
      dt.map <- cpg2hgnc[cpg_site %in% dt[type == x]$id]
      colnames(dt.map) <- c('id', 'hgnc')
      return(merge(dt[type == x], dt.map,  by = 'id'))
    } else if (x == 'cnv') {
      dt.map <- dt[type == x]
      dt.map$hgnc <- gsub("^.+?\\.", "", dt.map$variable)
      return(dt.map)
    } else if (x == 'gex' || x == 'mut') {
      dt.map <- ens2hgnc[ref_gene_id %in% dt[type == x]$id]
      colnames(dt.map) <- c('id', 'hgnc')
      return(merge(dt[type == x], dt.map,  by = 'id'))
    }
  }))
}
```

```{r pancancer_prediction, fig.height=9, fig.width=12}
# In this section, we make figures about pan-cancer classification 
# and some benchmarks of cohort-based unsupervised clustering across cancers

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.clinical_factors.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 8.27, height = 11.69, units = 'in')
Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type. B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs."

@@

```{r}
# Analysis of non-small-cell lung cancer types (LUSC vs LUAD)
# reference study: https://www.nature.com/articles/s41598-020-77284-8
M <- LFs$maui_factors$nsclc.cnv_gex_meth_mut
M <- scale(M) 
ct <- gsub("TCGA-", "", surv[match(rownames(M), bcr_patient_barcode)]$project)

# find type-specific LFs
subtype_lfs <- get_factor_specific_variables(M, ct)
top_vars <- get_top_by_label(subtype_lfs[padj < 0.05][order(padj)], 'ref_cl', 1)

@@

# top lf for LUSC
top_lf_lusc <- top_vars[ref_cl == 'LUSC']$variable
top_lf_luad <- top_vars[ref_cl == 'LUAD']$variable
df <- cbind(df, 
            M[,top_lf_lusc,drop = F],
            M[,top_lf_luad,drop = F])
p <- ggplot(df, aes(x = tSNE1, y = tSNE2)) 
p2 <- p + geom_point(aes(color = group)) + 
  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
            aes(x,y,label = text.label), size = 5)
p3 <- p + geom_point(aes_string(color = top_lf_lusc)) + 
  scale_color_gradient(low = 'gray', high = 'red') +
  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
            aes(x,y,label = text.label), size = 5)
p4 <- p + geom_point(aes_string(color = top_lf_luad)) + 
  scale_color_gradient(low = 'gray', high = 'red') +
  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
            aes(x,y,label = text.label), size = 5)

# detect subtype specific features that contribute most 
# to top LFs and make a decision tree plot  
## get top 10 markers per top LFs per group
top_vars <- get_top_by_label(subtype_lfs[padj < 0.05][order(padj)], 'ref_cl', 5)
W <- FWs$maui_factors$nsclc.cnv_gex_meth_mut
markers <- unlist(lapply(unique(top_vars$variable), function(x) {
  # get top features per lf and concatenate
  head(sort(abs(W[,x]), decreasing = T), 10)
}))

# import assay data and subset for markers 
A <- data.table::fread(file.path(settings$assay_output$folder, 
                                 'nsclc.cnv_gex_meth_mut.csv'))
A.sub <- A[V1 %in% names(markers)]
#convert to matrix
A.subM <- as.matrix(data.frame(A.sub[,-1],
                               row.names = A.sub$V1, 
                               check.names = F))
labels <- gsub("TCGA-", "", surv[match(colnames(A.subM), bcr_patient_barcode)]$project)

# further filter top features 
top_features <- get_factor_specific_variables(t(A.subM), labels)
# get top 5 per subtype
top_markers <- get_top_by_label(top_features[padj < 0.05][order(padj)], 
                                'ref_cl', topN = 5)
# map ids 
top_markers <- merge(top_markers, map_ids(top_markers$variable, cpg2hgnc, ens2hgnc), by = 'variable')

# use treeheatr and partykit package to compute/visualize decision tree
df <- data.frame(t(round(A.subM[top_markers$variable,], 3)))
colnames(df) <- paste(top_markers$type, top_markers$hgnc, sep = ":")
df$subtype <- as.factor(labels)

set.seed(123)
inTrain <- caret::createDataPartition(df$subtype, p = 0.6)
df.train <- df[inTrain[[1]],]
df.test <- df[-inTrain[[1]],]
p5 <- treeheatr::heat_tree(
  x = partykit::ctree(subtype ~ ., data = df.train),
  data_test = df.test,
  heat_rel_height = 0.4
)

# compute classification accuracy of using top 10 markers in a random forest model
rf_stats <- predict_cluster_labels(t(A.subM[top_markers$variable,]), labels, partition = 0.6)

Fig <-  cowplot::plot_grid(cowplot::plot_grid(p1, p2, p3, p4,
                         labels = 'AUTO', nrow = 2), 
                         p5, nrow = 2, labels = c('', 'E'))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.luad_vs_lusc.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       Fig, units = 'in', 
       width = 8.27, height = 11.69)

Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI latent factors.
A) ROC curves with AUC values on testing data (40% of the dataset) using an Elastic Net model with 5-fold cross-validation. B) t-SNE plot of MAUI LFs colored by cancer types. C) t-SNE plot of MAUI LFs colored by 
the top most-predictive latent factor for TCGA-LUSC samples. D) t-SNE plot of MAUI LFs colored by the top 
most predictive latent factor for TCGA-LUAD samples. E) Decision tree visualisation of top LF-associated omics markers detected that can be used to distinguish LUSC and LUAD subtypes."
```
